Titeln på avhandlingens språk

### 2022-10-04 model.tex
nice table

Finally, in \cref{alg:timings} we present a step-by-step breakdown of how this data is subsequently used to generate realistic execution times.
In plain English, a model which uses this processed data to generate execution times needs to maintain variables for duration, previous impairment, and most recent transition.
At the beginning of each step, the model
\begin{enumerate}
    \item Checks whether any change in impairment has occurred.
          If it has, it resets duration to \num{1} and updates the transition variable accordingly.
          If not, it increases the duration variable by one.
        %   unless that increase causes it to exceed the threshold of steps required to forget a transition.
        %   If that's the case, instead the model resets duration to \num{1} and sets the transition variable to ``no transition''.
    \item Discretizes the updated duration value.
    \item Finds all samples with binned neuroticism, impairment and duration matching the state of the system.
    If, and only if, the current duration falls within the \emph{short} duration bin, this data is further bisected according to the current value of the transition variable.
    \item Randomly samples the filtered data to obtain a realistic execution time value.
          \begin{itemize}
              \item Alternatively, a distribution can be first be fitted to the filtered samples.
                    The model then samples the theoretical distribution instead of the empirical samples.
          \end{itemize}
    \item Finally, updates the stored impairment value with the new impairment and proceeds to the next step.
\end{enumerate}

\begin{algorithm}
    \todo[inline]{Update algorithm.}
    \caption{Generating realistic execution times}\label{alg:timings}

    \SetKwInput{KwState}{Initial State}

    \KwIn{$\tau$: \ac{TTF} of the previous step.}
    \KwOut{$t_\text{exec}$: next execution time.}
    \KwState{%
        \begin{itemize}
            \item Binned neuroticism $N$.
            \item Number of steps to remember transitions for: $\delta \leftarrow 8$.
            \item Duration counter: $d \leftarrow 0$.
            \item Previous impairment: $I' \leftarrow \varnothing$.
            \item $B_\text{TTF}$ and $B_\text{duration}$ are functions for binning \ac{TTF} and duration into discrete levels, respectively.
            \item $F_\text{data}$ is a function which filters the data and returns only execution time samples which match the given levels of binned neuroticism, binned impairment, binned duration, and transition label.
        \end{itemize}
    }

    $I \leftarrow B_\text{TTF}(\tau)$ \tcc*{discretize \ac{TTF}}

    \uIf{($I' == \varnothing$) {\normalfont\textbf{or}}\\\hspace{1em}($T \neq \text{``NO TRANSITION''}$ {\normalfont\textbf{and}} $d + 1 > \delta$)}{%
        $d \leftarrow 1$\;
        $T \leftarrow \text{``NO TRANSITION''}$\;
    } \uElseIf{$I' < I$}{%
        $d \leftarrow 1$\;
        $T \leftarrow \text{``LOWER TO HIGHER''} $
    } \uElseIf{$I' > I$}{%
        $d \leftarrow 1$\;
        $T \leftarrow \text{``HIGHER TO LOWER''} $
    } \Else {%
        $d \leftarrow d + 1$\;
    }
    $I' \leftarrow I$ \tcc*{store impairment for later}
    $D \leftarrow B_\text{duration}(d)$ \tcc*{discretize duration}

    \tcc{next, filter the data to select only the matching samples}
    $E \leftarrow F_\text{data}(N, I, D, T)$ \;
    \tcc{finally, sample the data to return a realistic execution time}
    $t_\text{exec} \leftarrow {SAMPLE}(E)$\;

\end{algorithm}

@@

We use Python~\num{3.10} to implement two variants of the above described model.
The first of these samples execution times directly from the empirical data from~\cite{olguinmunoz:impact2021}; we will refer to it hereafter as the \emph{empirical model}.
The second corresponds to one which first fits a theoretical \ac{exGaussian} distribution to the data before sampling; we will refer to this one as the \emph{theoretical model}.
We also implement a reference model which uses a constant execution time of \SI{5.23}{\second}, which we in turn call the \emph{naive model}.
This value corresponds to the mean execution time of steps with the lowest level of impairment and a transition tag of \emph{no transition}.
\todo[inline]{Transition fade distance was 4 instead of 8, need to mention?}

>>>>

Finally, we present a step-by-step breakdown of how this data is subsequently used to generate realistic execution times.
A model which uses this processed data to generate execution times needs to maintain variables for duration, previous impairment, and most recent transition.
At the beginning of each step, it:
\begin{enumerate}
    \item Checks whether any change in impairment has occurred.
          If it has, it resets duration to \num{1} and updates the transition variable accordingly.
          If not, it increases the duration variable by one.
        %   unless that increase causes it to exceed the threshold of steps required to forget a transition.
        %   If that's the case, instead the model resets duration to \num{1} and sets the transition variable to ``no transition''.
    \item Discretizes the updated duration value.
    \item Finds all samples with binned neuroticism, impairment and duration matching the state of the system.
    If, and only if, the current duration falls within the \emph{short} duration bin, this data is further bisected according to the current value of the transition variable.
    \item Randomly samples the filtered data to obtain a realistic execution time value.
          \begin{itemize}
              \item Alternatively, a distribution can be first be fitted to the filtered samples.
                    The model then samples the theoretical distribution instead of the empirical samples.
          \end{itemize}
    \item Finally, updates the stored impairment value with the new impairment and proceeds to the next step.
\end{enumerate}

% \begin{algorithm}
%     \todo[inline]{Update algorithm.}
%     \caption{Generating realistic execution times}\label{alg:timings}

%     \SetKwInput{KwState}{Initial State}

%     \KwIn{$\tau$: \ac{TTF} of the previous step.}
%     \KwOut{$t_\text{exec}$: next execution time.}
%     \KwState{%
%         \begin{itemize}
%             \item Binned neuroticism $N$.
%             % \item Number of steps to remember transitions for: $\delta \leftarrow 8$.
%             \item Duration counter: $d \leftarrow 0$.
%             \item Previous impairment: $I' \leftarrow \varnothing$.
%             \item $B_\text{TTF}$ and $B_\text{duration}$ are functions for binning \ac{TTF} and duration into discrete levels, respectively.
%             \item $F_\text{data}$ is a function which filters the data and returns only execution time samples which match the given levels of binned neuroticism, binned impairment, binned duration, and transition label.
%         \end{itemize}
%     }

%     $I \leftarrow B_\text{TTF}(\tau)$ \tcc*{discretize \ac{TTF}}

%     \uIf{($I' == \varnothing$) {\normalfont\textbf{or}}\\\hspace{1em}($T \neq \text{``NO TRANSITION''}$ {\normalfont\textbf{and}} $d + 1 > \delta$)}{%
%         $d \leftarrow 1$\;
%         $T \leftarrow \text{``NO TRANSITION''}$\;
%     } \uElseIf{$I' < I$}{%
%         $d \leftarrow 1$\;
%         $T \leftarrow \text{``LOWER TO HIGHER''} $
%     } \uElseIf{$I' > I$}{%
%         $d \leftarrow 1$\;
%         $T \leftarrow \text{``HIGHER TO LOWER''} $
%     } \Else {%
%         $d \leftarrow d + 1$\;
%     }
%     $I' \leftarrow I$ \tcc*{store impairment for later}
%     $D \leftarrow B_\text{duration}(d)$ \tcc*{discretize duration}

%     \tcc{next, filter the data to select only the matching samples}
%     $E \leftarrow F_\text{data}(N, I, D, T)$ \;
%     \tcc{finally, sample the data to return a realistic execution time}
%     $t_\text{exec} \leftarrow {SAMPLE}(E)$\;

% \end{algorithm}

@@

We use Python~\num{3.10} to implement two variants of the above described timing model.
The first of these samples execution times directly from the empirical data from~\cite{olguinmunoz:impact2021}; we will refer to it hereafter as the \emph{empirical model}.
The second corresponds to one which first fits a theoretical \ac{exGaussian} distribution to the data before sampling; we will refer to this one as the \emph{theoretical model}.
We also implement three simpler reference timing models:
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
    \item a model which returns a constant execution time, corresponding to the mean execution time of all samples associated with the lowest impairment and the shortest duration
    % \item a model which randomly samples from the raw data without any grouping
    \item a model which fits an \ac{exGaussian} distribution to the raw data, without any grouping, and then randomly samples from it  
\end{enumerate*}.
We will refer to these models as the \emph{constant} and \emph{fitted-naive} model, respectively.

\todo[inline]{Insert synthetic, controlled setup here.}

\begin{figure*}
    \centering
    \includegraphics[width=.9\textwidth]{figs/model_comp_controlled_ttf.png}
    \begin{tabular}{ll >{\columncolor[HTML]{EAEAEA}}r >{\columncolor[HTML]{EAEAEA}}r rr}
    \toprule
    {} & {} & \multicolumn{2}{>{\columncolor[HTML]{EAEAEA}}c}{\ac{TTF} \( = \) \SI{0}{\second}} & \multicolumn{2}{c}{\ac{TTF} \( = \) \SI{5}{\second}} \\
    {Model} & {Neuroticism} & {Total Duration} & {Diff.\ w.r.t.\ fitted-naive} & {Total Duration} & {Diff.\ w.r.t.\ fitted-naive} \\
    \midrule
    fitted-naive & n/a & \SI{576.74}{\second} & & \SI{1084.57}{\second} & \\
    \multirow[c]{2}{*}{empirical} & low & \SI{511.18}{\second} & \SI{-11.37}{\percent} & \SI{1083.40}{\second} & \SI{-0.11}{\percent} \\
    & high & \SI{533.29}{\second} & \SI{-7.53}{\percent} & \SI{1245.46}{\second} & +\SI{14.84}{\percent} \\
    \multirow[c]{2}{*}{theoretical} & low & \SI{504.73}{\second} & \SI{-12.49}{\percent} & \SI{1080.16}{\second} & \SI{-0.41}{\percent} \\
    & high & \SI{539.00}{\second} & \SI{-6.54}{\percent} & \SI{1225.68}{\second} & +\SI{13.01}{\percent} \\
    \bottomrule
    \end{tabular}
    \caption{
        Comparison of the total duration of \num{100}-step tasks for different models in the controlled \ac{TTF} setup.
        The results are separated for clarity, with the top plot showing the results for the realistic models, and the bottom one for the reference models.
    }\label{fig:ctrlttf}
\end{figure*}

We first compare these models using a controlled, ideal setup with \num{25} constant, linearly distributed \acp{TTF} in the \SIrange[]{0}{5}{\second} range.
For each \ac{TTF}, the models execute \num{30} independent repetitions of a \num{100} step task.

We show the result of these runs in \cref{fig:ctrlttf}.
At the highest level of neuroticism, the realistic models result in \textasciitilde{}\SI{14}{\percent} longer total durations when compared to the \emph{fitted-naive} reference model.
